---
title: "Mastering NumPy and Base R"

author: "Geethma Jayasinghe"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    number-sections: true
    code-copy: true
    code-fold: show
---

# The Basics of NumPy Arrays

Data manipulation in Python is nearly synonymous with NumPy array manipulation: even newer tools like Pandas are built around the NumPy array. This section will present several examples using NumPy array manipulation.
We'll cover a few categories of basic array manipulations here:

*Attributes of arrays*
:   Determining the size, shape, memory consumption, and data types of arrays

*Indexing of arrays*
:   Getting and setting the value of individual array elements

*Slicing of arrays*
:   Getting and setting smaller subarrays within a larger array

*Reshaping of arrays*
:   Changing the shape of a given array

*Joining and splitting of arrays*
:   Combining multiple arrays into one, and splitting one array into many

---

# NumPy Array Attributes

First let's discuss some useful array attributes.

::: {.panel-tabset}

## NumPy (Python)

```{python}
import numpy as np

np.random.seed(0)

x1 = np.random.randint(10, size=6)
x2 = np.random.randint(10, size=(3, 4))
x3 = np.random.randint(10, size=(3, 4, 5))
```

Each array has attributes `ndim`, `shape`, and `size`.

```{python}
print("x3 ndim:", x3.ndim)
print("x3 shape:", x3.shape)
print("x3 size:", x3.size)
```

### What Happened?

- `ndim` — number of dimensions
- `shape` — size of each dimension
- `size` — total number of elements
- `dtype` — data type of elements
- `itemsize` — bytes per element
- `nbytes` — total memory used

## Base R

```{r}
set.seed(0)

x1 <- sample(0:9, 6, replace = TRUE)
x2 <- matrix(sample(0:9, 12, replace = TRUE),
             nrow = 3,
             ncol = 4)

x3 <- array(sample(0:9, 60, replace = TRUE),
            dim = c(3, 4, 5))
```

```{r}
cat("ndim:", length(dim(x3)), "\n")
cat("shape:", dim(x3), "\n")
cat("size:", length(x3), "\n")
```

:::
# Array Indexing: Accessing Single Elements

If you are familiar with Python's standard list indexing, indexing in NumPy will feel quite familiar. In a one-dimensional array, you can access the $i$th value (counting from zero) by specifying the desired index in square brackets, just as with Python lists.

::: {.panel-tabset}

## NumPy (Python)

```{python}
x1
```

```{python}
x1[0]
```

```{python}
x1[4]
```


## Base R

```{r}
x1[1]   # first element (R is 1-based!)
```

```{r}
x1[5]   # fifth element
```

```{r}
x1[length(x1)]  # last element
```
:::

---

# Array Slicing: Accessing Subarrays

Just as we can use square brackets to access individual array elements, we can also use them to access subarrays with the *slice* notation, marked by the colon (`:`) character. The NumPy slicing syntax follows that of the standard Python list; to access a slice of an array `x`, use this:

```
x[start:stop:step]
```

If any of these are unspecified, they default to the values `start=0`, `stop=size of dimension`, `step=1`. We'll take a look at accessing subarrays in one dimension and in multiple dimensions.

## One-dimensional subarrays

::: {.panel-tabset}

## NumPy (Python)

```{python}
x = np.arange(10)
x
```

```{python}
x[:5]   # first five elements
```

```{python}
x[5:]   # elements after index 5
```

```{python}
x[4:7]  # middle subarray
```

```{python}
x[::2]  # every other element
```

## Base R

```{r}
x <- 0:9
x
```

```{r}
x[1:5]               # first five elements (1-based)
```

```{r}
x[6:length(x)]       # elements after position 5
```

```{r}
x[5:7]               # middle subarray
```

```{r}
x[seq(1, 10, by = 2)]  # every other element
```

```{r}
rev(x)               # all elements, reversed
```

:::

## Multidimensional subarrays

Multidimensional slices work in the same way, with multiple slices separated by commas.

::: {.panel-tabset}

## NumPy (Python)

```{python}
x2
```

```{python}
x2[:2, :3]   # two rows, three columns
```


## Base R

```{r}
x2[1:2, 1:3]   # two rows, three columns
```

```{r}
x2[, 1]        # first column (all rows)
```

```{r}
x2[1, ]        # first row (all columns)
```

:::

---

---

# Reshaping of Arrays

Another useful type of operation is reshaping of arrays. The most flexible way of doing this is with the `reshape()` method. For example, if you want to put the numbers 1 through 9 in a 3×3 grid, you can do the following.

Note that for this to work, the size of the initial array must match the size of the reshaped array. Where possible, the `reshape` method will use a no-copy view of the initial array, but with noncontiguous memory buffers this is not always the case.

::: {.panel-tabset}

## NumPy (Python)

```{python}
grid = np.arange(1, 10).reshape((3, 3))
print(grid)
```

### What Happened?

- `reshape((3,3))` converts a 1-D array into a 3×3 matrix.
- The total number of elements must remain unchanged.
- NumPy attempts to avoid copying data whenever possible.

## Base R

```{r}
grid <- matrix(1:9, nrow = 3, byrow = TRUE)
print(grid)
```

### R Equivalent

- `matrix()` is the most common way to reshape vectors into matrices.
- The total number of elements must match the target dimensions.

:::

---

# Array Concatenation

All of the preceding routines worked on single arrays. It's also possible to combine multiple arrays into one, and to conversely split a single array into multiple arrays.

## Concatenation of arrays

Concatenation, or joining of two arrays in NumPy, is primarily accomplished through the routines `np.concatenate`, `np.vstack`, and `np.hstack`. `np.concatenate` takes a tuple or list of arrays as its first argument.

::: {.panel-tabset}

## NumPy (Python)

```{python}
x = np.array([1, 2, 3])
y = np.array([3, 2, 1])
np.concatenate([x, y])
```

You can also concatenate more than two arrays at once:

```{python}
z = [99, 99, 99]
print(np.concatenate([x, y, z]))
```

`np.concatenate` can also be used for two-dimensional arrays:

```{python}
grid = np.array([[1, 2, 3],
                 [4, 5, 6]])

np.concatenate([grid, grid])
```

## Base R

```{r}
x <- c(1, 2, 3)
y <- c(3, 2, 1)

c(x, y)
```

```{r}
z <- c(99, 99, 99)

c(x, y, z)
```

```{r}
grid <- matrix(c(1, 2, 3,
                 4, 5, 6),
               nrow = 2,
               byrow = TRUE)

rbind(grid, grid)
```

```{r}
cbind(grid, grid)
```

:::