---
title: "Computation on NumPy Arrays"
author: N.G.S. Minimuthu
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    number-sections: true
    code-copy: true
    code-fold: show
---

# Universal Functions (UFuncs)

Computation on NumPy arrays can be very fast, or it can be very slow. The key to making it fast is to use **vectorized operations** — applying an operation to the whole array at once instead of looping through each element.

> These vectorized operations are implemented through NumPy's **universal functions**. Any time you see a loop over an array in Python, think about whether a ufunc can replace it.


---

# Array Arithmetic

NumPy uses Python's standard arithmetic operators directly on arrays. Each operator applies to every element at once.

::: {.panel-tabset}

## NumPy (Python)

```{python}
import numpy as np

x = np.arange(4)
print("x      =", x)
print("x + 5  =", x + 5)
print("x - 2  =", x - 2)
print("x * 3  =", x * 3)
print("x / 2  =", x / 2)
print("x // 2 =", x // 2)   # floor division
print("x ** 2 =", x ** 2)   # exponentiation
print("x % 2  =", x % 2)    # modulus
```



- `//` is floor division (drops the decimal), `**` is power, `%` is the remainder.

## Base R

```{r}
x <- 0:3
x <- 0:3
print(x + 5)
print(x - 2)
print(x * 3)
print(x / 2)
print(x %/% 2)  # floor division
print(x ^ 2)     # exponentiation
print(x %% 2)    # modulus
```


- The operators are nearly identical, with minor syntax differences: `^` instead of `**`, `%/%` instead of `//`.

:::

---

# Useful Math Functions

Beyond basic arithmetic, NumPy provides functions for absolute values, trigonometry, exponentials, and logarithms — all applied element-wise.

::: {.panel-tabset}

## NumPy (Python)

```{python}
import numpy as np

# Absolute value
x = np.array([-3, -1, 0, 1, 3])
print("abs:", np.abs(x))

# Trigonometry
theta = np.linspace(0, np.pi, 3)
print("sin:", np.sin(theta).round(2))
print("cos:", np.cos(theta).round(2))

# Exponentials and logarithms
x = np.array([1, 2, 3])
print("e^x:    ", np.exp(x).round(3))
print("ln(x):  ", np.log(x).round(3))
print("log2(x):", np.log2(x).round(3))
```


## Base R

```{r}
# Absolute value
x <- c(-3, -1, 0, 1, 3)
print(abs(x))

# Trigonometry
theta <- seq(0, pi, length.out = 3)
print(round(sin(theta), 2))
print(round(cos(theta), 2))

# Exponentials and logarithms
x <- c(1, 2, 3)
print(round(exp(x), 3))
print(round(log(x), 3))
print(round(log2(x), 3))
```

### What Happened?

- R's `abs()`, `sin()`, `cos()`, `exp()`, `log()` and `log2()` work exactly the same way.
- The function names are almost identical.

:::

---

# Aggregations

After doing computations, you often want to summarise an array into a single value — the sum, mean, min, or max.

::: {.panel-tabset}

## NumPy (Python)

```{python}
import numpy as np

np.random.seed(0)
arr = np.random.randint(1, 100, size=6)
print("Array:", arr)

print("sum:   ", arr.sum())
print("min:   ", arr.min())
print("max:   ", arr.max())
print("mean:  ", arr.mean().round(2))
print("std:   ", arr.std().round(2))
print("median:", np.median(arr))
```


- You can call them as `np.sum(arr)` or as a method directly on the array: `arr.sum()` — both work.
- Always prefer `arr.sum()` over Python's built-in `sum(arr)` on NumPy arrays — the NumPy version is much faster.

## Base R

```{r}
set.seed(0)
arr <- sample(1:99, 6, replace = TRUE)
cat("Array:", arr, "\n")

print(arr)
print(sum(arr))
print(min(arr))
print(max(arr))
print(round(mean(arr), 2))
print(round(sd(arr), 2))
print(median(arr))
```

### What Happened?

- R's aggregation functions are identical in concept: `sum()`, `min()`, `max()`, `mean()`, `sd()`, `median()`.
- The main difference is that in NumPy you can call them as methods on the array itself (`arr.mean()`), which R does not do.

:::


---

# NumPy vs R: Computation Quick Reference

| Task | NumPy (Python) | Base R |
|------|---------------|--------|
| Add scalar | `x + 5` | `x + 5` |
| Power | `x ** 2` | `x ^ 2` |
| Floor division | `x // 2` | `x %/% 2` |
| Modulus | `x % 2` | `x %% 2` |
| Absolute value | `np.abs(x)` | `abs(x)` |
| Sine / Cosine | `np.sin(x)` / `np.cos(x)` | `sin(x)` / `cos(x)` |
| Natural log | `np.log(x)` | `log(x)` |
| Exponential | `np.exp(x)` | `exp(x)` |
| Sum | `arr.sum()` | `sum(arr)` |
| Min / Max | `arr.min()` / `arr.max()` | `min(arr)` / `max(arr)` |
| Mean | `arr.mean()` | `mean(arr)` |
| Standard deviation | `arr.std()` | `sd(arr)` |
| Median | `np.median(arr)` | `median(arr)` |


> If you already know R, NumPy computation will feel very familiar. The operators and function names are almost the same — the biggest difference is that in NumPy you can call aggregations as methods directly on the array (`arr.mean()`).


