---
title: "Time Series Analysis Walkthrough: Python vs R"
subtitle: "Working Through AirPassengers, Step by Step, in Both Languages"
author: "G. I. Hansamal"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    fig-width: 8
    fig-height: 4.5
engine: knitr
execute:
  warning: false
---

## About This Walkthrough

We're going to work through a complete time series analysis twice — once in
Python, once in R — using the same dataset and the same modeling decisions
both times. Each step below has a tabbed panel, so you can click between
"Python" and "R" and see how the same idea is written in each language.

Let's use the **AirPassengers** dataset for this: monthly totals of
international airline passengers from 1949 to 1960. It's a familiar one in
time series courses, but that's actually the point here — it has a clean
upward trend, a yearly seasonal pattern, and the seasonal swings get bigger as
the numbers get bigger. We'll lean on that last property a few times as we go,
to motivate why we make certain modeling choices instead of just asserting
them.

::: {.callout-note}
**Before you render this**

- **R:** needs the `tseries` package (for `adf.test()` / `kpss.test()`) and `reticulate` (to run the Python chunks). Everything else on the R side — `decompose()`, `arima()`, `acf()`/`pacf()` — ships with base R.
- **Python:** needs `numpy`, `pandas`, `matplotlib`, `scipy`, and `statsmodels` available to whatever Python `reticulate` points at.
- We pull the dataset into Python using `statsmodels`' `get_rdataset()`, which downloads it from the same public Rdatasets archive R's built-in copy traces back to. So the first render needs internet access; after that it's cached locally (`cache=True`), so we don't re-download it every time.
:::

## Step 1: Load the Data and Set Up a Train/Test Split

Let's start by loading the data. R already has `AirPassengers` built in. On
the Python side, `statsmodels` can fetch the exact same dataset by name
through `get_rdataset()` — so we're not approximating it with something
similar, it's the same 144 monthly values either way.

We'll also hold out the last 12 months (all of 1960) right away, before we do
anything else. Everything we build later — the decomposition forecast and the
ARIMA forecast — gets checked against those 12 real, held-back values instead
of just being eyeballed on a chart.

::: {.panel-tabset}

### Python

```{python}
import warnings
warnings.filterwarnings("ignore")

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
import statsmodels.api as sm

from statsmodels.tsa.seasonal import seasonal_decompose
from statsmodels.tsa.stattools import adfuller, kpss
from statsmodels.stats.diagnostic import acorr_ljungbox
from statsmodels.graphics.tsaplots import plot_acf, plot_pacf
from statsmodels.tsa.arima.model import ARIMA

raw = sm.datasets.get_rdataset("AirPassengers", "datasets", cache=True).data
dates = pd.date_range("1949-01-01", periods=len(raw), freq="MS")
ts_data = pd.Series(raw["value"].values, index=dates, name="passengers")

h = 12
train = ts_data.iloc[:-h]
test = ts_data.iloc[-h:]

print(ts_data.head())
print(f"\nTotal months: {len(ts_data)}  |  Train: {len(train)}  |  Test: {len(test)}")
```

### R

```{r}
# install.packages("tseries")  # run once if you don't already have it
library(tseries)

data("AirPassengers")

h <- 12
train <- window(AirPassengers, end = c(1959, 12))
test  <- window(AirPassengers, start = c(1960, 1))

head(AirPassengers)
cat("\nTotal months:", length(AirPassengers),
    " | Train:", length(train), " | Test:", length(test), "\n")
```

:::

Both tabs should print the same first six values (112, 118, 132, 129, 121, 135
thousand passengers) and the same 132/12 train/test split. If those don't
match, we've got a data-loading problem to fix before going any further.

## Step 2: Plot the Time Series

Let's just look at the data before we do anything to it.

::: {.panel-tabset}

### Python

```{python}
plt.figure(figsize=(9, 4))
plt.plot(ts_data.index, ts_data.values, color="steelblue")
plt.title("AirPassengers: Monthly International Passengers (1949-1960)")
plt.xlabel("Date")
plt.ylabel("Passengers (thousands)")
plt.tight_layout()
plt.show()
```

### R

```{r}
plot(AirPassengers,
     main = "AirPassengers: Monthly International Passengers (1949-1960)",
     xlab = "Year", ylab = "Passengers (thousands)", col = "steelblue")
```

:::

Already we can see most of what we need: a steady upward trend, a repeating
yearly pattern (summer peaks, winter dips), and — look closely — the size of
those yearly peaks and dips grows as we move right. Early on, the swing within
a year is maybe 20-30 thousand passengers; by the end it's closer to 100. We'll
come back to that widening pattern in the next step, since it's the visual
clue for choosing a **multiplicative** rather than purely **additive**
seasonal model.

## Step 3: Classical Decomposition — Trend, Seasonal Index, and Residual

Now let's break the series down into its parts. Classical decomposition
splits it into trend, seasonal, and residual pieces, combined one of two ways:

- **Additive:** `observed = trend + seasonal + residual`
- **Multiplicative:** `observed = trend × seasonal × residual`

We'll fit both on the **training** data and compare them — and instead of
just plotting the pieces, let's also pull out the actual numbers: the trend
line's values and the twelve seasonal indices (one per calendar month).

::: {.panel-tabset}

### Python

```{python}
add_decomp = seasonal_decompose(train, model="additive", period=12)
fig = add_decomp.plot()
fig.set_size_inches(9, 6)
fig.suptitle("Additive Decomposition")
plt.tight_layout()
plt.show()

mul_decomp = seasonal_decompose(train, model="multiplicative", period=12)
fig = mul_decomp.plot()
fig.set_size_inches(9, 6)
fig.suptitle("Multiplicative Decomposition")
plt.tight_layout()
plt.show()

# Pull the trend line out as numbers, not just a picture
print("Trend component, first 3 estimated months:")
print(add_decomp.trend.dropna().head(3).round(2))
print("\nTrend component, last 3 estimated months:")
print(add_decomp.trend.dropna().tail(3).round(2))

# The seasonal component repeats every 12 months -- the first 12 values
# ARE the seasonal index, one per calendar month
month_names = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]

add_seasonal_index = add_decomp.seasonal.iloc[:12]
add_seasonal_index.index = month_names
print("\nAdditive seasonal index (thousands of passengers, +/- around trend):")
print(add_seasonal_index.round(2))

mul_seasonal_index = mul_decomp.seasonal.iloc[:12]
mul_seasonal_index.index = month_names
print("\nMultiplicative seasonal index (ratio relative to trend):")
print(mul_seasonal_index.round(3))

print(f"\nAdditive residual std dev:        {np.nanstd(add_decomp.resid):.2f} (thousands of passengers)")
print(f"Multiplicative residual std dev:  {np.nanstd(mul_decomp.resid):.4f} (ratio, centered on 1.0)")
```

### R

```{r}
decomp_add <- decompose(train, type = "additive")
plot(decomp_add)

decomp_mul <- decompose(train, type = "multiplicative")
plot(decomp_mul)

# Pull the trend line out as numbers, not just a picture
cat("Trend component, first 3 estimated months:\n")
print(round(head(na.omit(decomp_add$trend), 3), 2))
cat("\nTrend component, last 3 estimated months:\n")
print(round(tail(na.omit(decomp_add$trend), 3), 2))

# decompose() already stores the one-cycle seasonal index in $figure
add_seasonal_index <- setNames(decomp_add$figure, month.abb)
cat("\nAdditive seasonal index (thousands of passengers, +/- around trend):\n")
print(round(add_seasonal_index, 2))

mul_seasonal_index <- setNames(decomp_mul$figure, month.abb)
cat("\nMultiplicative seasonal index (ratio relative to trend):\n")
print(round(mul_seasonal_index, 3))

cat("\nAdditive residual std dev:       ", round(sd(na.omit(decomp_add$random)), 2),
    "(thousands of passengers)\n")
cat("Multiplicative residual std dev: ", round(sd(na.omit(decomp_mul$random)), 4),
    "(ratio, centered on 1.0)\n")
```

:::

A few things worth reading off that output rather than just the plots: the
**additive seasonal index** tells us, in thousands of passengers, how far
above or below trend each calendar month typically runs — July and August are
the strongest (around +55 to +60), February and November the weakest (around
-30 to -50). The **multiplicative seasonal index** tells the same story as a
ratio: July runs about 22% above trend, November about 20% below.

Now compare the residual spread between the two. In the additive version, the
leftover noise visibly fans out — small in the early years, noticeably bigger
toward the end — meaning the additive model isn't fully capturing the
seasonal pattern. In the multiplicative version, the residual (a ratio
centered on 1.0) stays small and fairly even across the whole period. That
matches what we noticed back in Step 2 — this dataset behaves
multiplicatively, and the printed standard deviations confirm it.

## Step 4: Statistical Tests for Stationarity

Before we fit any AR/MA-style model, let's check whether the series is
(close to) **stationary** — roughly constant mean and variance over time.
Three checks:

- **Augmented Dickey-Fuller (ADF):** null hypothesis is "non-stationary." A small p-value lets us reject that, i.e. it argues *for* stationarity.
- **KPSS:** the opposite null — "stationary." A small p-value here argues *against* stationarity. Running both is a useful cross-check, since they can occasionally disagree.
- **Ljung-Box:** tests whether there's significant leftover autocorrelation at a given lag — useful here, and again later on model residuals.

::: {.panel-tabset}

### Python

```{python}
adf_raw = adfuller(train)
print(f"ADF on raw series  -> statistic={adf_raw[0]:.3f}, p-value={adf_raw[1]:.3f}")

kpss_raw = kpss(train, regression="c")
print(f"KPSS on raw series -> statistic={kpss_raw[0]:.3f}, p-value={kpss_raw[1]:.3f}")

lb_raw = acorr_ljungbox(train, lags=[12], return_df=True)
print("\nLjung-Box test (lag 12) on raw series:")
print(lb_raw)

# Stabilize the variance with a log transform, then remove the trend with one difference
log_train = np.log(train)
log_diff = log_train.diff().dropna()

adf_diff = adfuller(log_diff)
print(f"\nADF on log + differenced series -> statistic={adf_diff[0]:.3f}, p-value={adf_diff[1]:.3f}")
```

### R

```{r}
adf_raw <- adf.test(train)
print(adf_raw)

kpss_raw <- kpss.test(train)
print(kpss_raw)

lb_raw <- Box.test(train, lag = 12, type = "Ljung-Box")
print(lb_raw)

# Stabilize the variance with a log transform, then remove the trend with one difference
log_train <- log(train)
log_diff <- diff(log_train)

adf_diff <- adf.test(log_diff)
print(adf_diff)
```

:::

On the raw training series, we'd expect the ADF p-value to come back large
(can't reject "non-stationary") and the KPSS p-value to come back small
(rejects "stationary") — both pointing the same direction, matching the
obvious trend in the plot. The Ljung-Box test should also come back with a
tiny p-value, confirming there's a lot of structure (trend + seasonality)
still left to account for.

The log transform we just applied addresses the growing seasonal swings from
Step 3 — working in log space turns a multiplicative pattern into an additive
one, which is exactly the kind of pattern ARIMA-family models assume. Taking
one difference on top of that removes the trend. After both steps, the ADF
p-value on `log_diff` should drop well below 0.05 — that series is stationary
enough to model directly. This is exactly what `ARIMA(p, 1, q)` is doing
internally once we hand it the logged series with `d = 1` in Step 7.

## Step 5: Forecast With Classical Decomposition

Let's see how far a simple, easy-to-explain forecast gets us, before we bring
in ARIMA at all:

1. Fit a straight line through the trend component (from the **training**
   decomposition only) and extend it 12 months forward.
2. Repeat the last full seasonal cycle for those same 12 months.
3. Add the two together.
4. Check the result against the actual held-out 1960 values, using three
   error measures: RMSE, MAE, and MAPE (mean absolute percentage error).

::: {.panel-tabset}

### Python

```{python}
trend_comp = add_decomp.trend.dropna()
t_trend = np.arange(len(trend_comp))
slope, intercept = np.polyfit(t_trend, trend_comp.values, 1)

t_future = np.arange(len(trend_comp), len(trend_comp) + h)
trend_forecast = slope * t_future + intercept

seasonal_cycle = add_decomp.seasonal.values[-12:]
seasonal_forecast = np.tile(seasonal_cycle, h // 12 + 1)[:h]

decomp_forecast = trend_forecast + seasonal_forecast

rmse_decomp = np.sqrt(np.mean((decomp_forecast - test.values) ** 2))
mae_decomp = np.mean(np.abs(decomp_forecast - test.values))
mape_decomp = np.mean(np.abs((test.values - decomp_forecast) / test.values)) * 100

print("Classical decomposition forecast vs. actual 1960:")
print(f"  RMSE = {rmse_decomp:.1f} (thousands)")
print(f"  MAE  = {mae_decomp:.1f} (thousands)")
print(f"  MAPE = {mape_decomp:.1f}%")

plt.figure(figsize=(9, 4))
plt.plot(train.index, train.values, label="Train")
plt.plot(test.index, test.values, label="Actual 1960", color="black", linestyle="--")
plt.plot(test.index, decomp_forecast, label="Decomposition Forecast", color="darkorange")
plt.title("Classical Decomposition Forecast vs. Actual")
plt.legend()
plt.tight_layout()
plt.show()
```

### R

```{r}
trend_comp <- na.omit(decomp_add$trend)
t_trend <- seq_along(trend_comp)
fit <- lm(trend_comp ~ t_trend)

t_future <- seq(length(trend_comp) + 1, length(trend_comp) + h)
trend_forecast <- predict(fit, newdata = data.frame(t_trend = t_future))

# decompose() stores one repeating seasonal cycle in $figure
seasonal_forecast <- rep(decomp_add$figure, length.out = h)

decomp_forecast <- trend_forecast + seasonal_forecast

rmse_decomp <- sqrt(mean((decomp_forecast - as.numeric(test))^2))
mae_decomp <- mean(abs(decomp_forecast - as.numeric(test)))
mape_decomp <- mean(abs((as.numeric(test) - decomp_forecast) / as.numeric(test))) * 100

cat("Classical decomposition forecast vs. actual 1960:\n")
cat("  RMSE =", round(rmse_decomp, 1), "(thousands)\n")
cat("  MAE  =", round(mae_decomp, 1), "(thousands)\n")
cat("  MAPE =", round(mape_decomp, 1), "%\n")

forecast_ts <- ts(decomp_forecast, start = c(1960, 1), frequency = 12)

ts.plot(train, test, forecast_ts,
        col = c("steelblue", "black", "darkorange"), lty = c(1, 2, 1),
        main = "Classical Decomposition Forecast vs. Actual")
legend("topleft", legend = c("Train", "Actual 1960", "Decomposition Forecast"),
       col = c("steelblue", "black", "darkorange"), lty = c(1, 2, 1))
```

:::

The RMSE and MAE are in the same units as the data (thousands of monthly
passengers), so they're directly readable — an MAE around 45-50 means the
forecast was off by roughly 45-50 thousand passengers per month on average
over 1960. MAPE gives us the same idea as a percentage, which is handy for
comparing across datasets later. Worth noticing on the chart too: because
this forecast used the *additive* decomposition's trend and seasonal figure,
it tends to undershoot the biggest summer peaks a little — the same weakness
we flagged for additive decomposition back in Step 3. Let's keep that number
in mind; we'll compare it directly to ARIMA's error in Step 9.

## Step 6: ACF and PACF — Choosing Model Orders

Now let's switch over to ARMA-family models. We already stationarized the
series in Step 4 (`log_diff` = one difference of the logged training data),
so we can go straight to the ACF and PACF on that. These plots are normally
how you'd pick starting values for `p` and `q`: bars that stick out past the
shaded/dashed confidence band are the lags worth paying attention to.

::: {.panel-tabset}

### Python

```{python}
fig, axes = plt.subplots(1, 2, figsize=(10, 4))
plot_acf(log_diff, lags=24, ax=axes[0])
plot_pacf(log_diff, lags=24, ax=axes[1])
axes[0].set_title("ACF (log, differenced)")
axes[1].set_title("PACF (log, differenced)")
plt.tight_layout()
plt.show()
```

### R

```{r}
par(mfrow = c(1, 2))
acf(log_diff, lag.max = 24, main = "ACF (log, differenced)")
pacf(log_diff, lag.max = 24, main = "PACF (log, differenced)")
par(mfrow = c(1, 1))
```

:::

The early lags (1-3) usually show up as significant in both plots, which is
typical of an ARMA-type process — that's why an order like (2, ., 2) is a
reasonable starting point in the next step. We'll also likely spot a spike
around lag 12 (maybe 24) — that's the seasonal cycle poking through. A single
ordinary difference removes the *trend* but not *seasonality*; that's a real
limitation of plain ARIMA on a series like this one, and it's worth
remembering once we get to the forecast comparison in Step 9.

## Step 7: Fit AR, MA, ARMA, and ARIMA Models

Let's fit all four model types — they're really all special cases of
`ARIMA(p, d, q)` — on the **logged** training data, for the variance-
stabilizing reason discussed in Step 4:

| Model         | (p, d, q) |
|---------------|-----------|
| AR(2)         | (2, 0, 0) |
| MA(2)         | (0, 0, 2) |
| ARMA(2,2)     | (2, 0, 2) |
| ARIMA(2,1,2)  | (2, 1, 2) |

We'll compare them two ways: **AIC** (lower is better — rewards a closer fit
but penalizes unnecessary extra parameters) and **in-sample RMSE** on the log
scale (how far the model's own fitted values were from the training data it
was fit on).

::: {.panel-tabset}

### Python

```{python}
orders = {
    "AR(2)": (2, 0, 0),
    "MA(2)": (0, 0, 2),
    "ARMA(2,2)": (2, 0, 2),
    "ARIMA(2,1,2)": (2, 1, 2),
}

rows = []
for name, order in orders.items():
    p, d, q = order
    fitted_model = ARIMA(log_train, order=order).fit()
    # skip the first d points -- differencing leaves them without a real fitted value
    resid = log_train.iloc[d:] - fitted_model.fittedvalues.iloc[d:]
    rmse_log = np.sqrt(np.mean(resid ** 2))
    rows.append({"model": name, "AIC": fitted_model.aic, "in_sample_RMSE_log": rmse_log})

model_comparison = pd.DataFrame(rows).set_index("model")
print(model_comparison.round(4))
```

### R

```{r}
orders <- list(
  "AR(2)"        = c(2, 0, 0),
  "MA(2)"        = c(0, 0, 2),
  "ARMA(2,2)"    = c(2, 0, 2),
  "ARIMA(2,1,2)" = c(2, 1, 2)
)

model_comparison <- t(sapply(orders, function(o) {
  fit <- arima(log_train, order = o)
  c(AIC = AIC(fit), in_sample_RMSE_log = sqrt(mean(fit$residuals^2)))
}))
print(round(model_comparison, 4))
```

:::

`ARIMA(2,1,2)` should come out clearly ahead of the other three on both AIC
*and* in-sample RMSE, and that's not a coincidence — AR(2) and MA(2) on their
own have no mechanism for handling a strong, sustained trend like this one;
only the `d = 1` differencing term lets the model follow a moving level.
That's the practical reason ARIMA tends to beat plain AR or MA models on
trending data, beyond just "it has more parameters." (One small note: Python
and R estimate these models slightly differently under the hood — conditional
sum-of-squares vs. a state-space/Kalman filter approach by default — so don't
expect the two tables to match to the decimal. They should agree on which
model wins, though.)

## Step 8: Forecast With ARIMA

Let's take `ARIMA(2,1,2)` — fit on the log scale — and forecast the same 12
held-out months. We'll exponentiate the forecast (and its confidence
interval) back to the original passenger-count scale before comparing it to
`test`, using the same three error measures as Step 5.

::: {.panel-tabset}

### Python

```{python}
final_model = ARIMA(log_train, order=(2, 1, 2)).fit()
fc = final_model.get_forecast(steps=h)

fc_mean = np.exp(fc.predicted_mean)
ci = np.exp(fc.conf_int())

rmse_arima = np.sqrt(np.mean((fc_mean.values - test.values) ** 2))
mae_arima = np.mean(np.abs(fc_mean.values - test.values))
mape_arima = np.mean(np.abs((test.values - fc_mean.values) / test.values)) * 100

print("ARIMA(2,1,2) forecast vs. actual 1960:")
print(f"  RMSE = {rmse_arima:.1f} (thousands)")
print(f"  MAE  = {mae_arima:.1f} (thousands)")
print(f"  MAPE = {mape_arima:.1f}%")

plt.figure(figsize=(9, 4))
plt.plot(train.index, train.values, label="Train")
plt.plot(test.index, test.values, label="Actual 1960", color="black", linestyle="--")
plt.plot(test.index, fc_mean.values, label="ARIMA Forecast", color="darkgreen")
plt.fill_between(test.index, ci.iloc[:, 0], ci.iloc[:, 1], color="darkgreen", alpha=0.2, label="95% CI")
plt.title("ARIMA(2,1,2) Forecast (back-transformed from log scale) vs. Actual")
plt.legend()
plt.tight_layout()
plt.show()
```

### R

```{r}
final_fit <- arima(log_train, order = c(2, 1, 2))
pred <- predict(final_fit, n.ahead = h)

upper_log <- pred$pred + 1.96 * pred$se
lower_log <- pred$pred - 1.96 * pred$se

fc_mean <- exp(pred$pred)
upper <- exp(upper_log)
lower <- exp(lower_log)

rmse_arima <- sqrt(mean((as.numeric(fc_mean) - as.numeric(test))^2))
mae_arima <- mean(abs(as.numeric(fc_mean) - as.numeric(test)))
mape_arima <- mean(abs((as.numeric(test) - as.numeric(fc_mean)) / as.numeric(test))) * 100

cat("ARIMA(2,1,2) forecast vs. actual 1960:\n")
cat("  RMSE =", round(rmse_arima, 1), "(thousands)\n")
cat("  MAE  =", round(mae_arima, 1), "(thousands)\n")
cat("  MAPE =", round(mape_arima, 1), "%\n")

forecast_ts <- ts(as.numeric(fc_mean), start = c(1960, 1), frequency = 12)
upper_ts <- ts(as.numeric(upper), start = c(1960, 1), frequency = 12)
lower_ts <- ts(as.numeric(lower), start = c(1960, 1), frequency = 12)

ts.plot(train, test, forecast_ts, upper_ts, lower_ts,
        col = c("steelblue", "black", "darkgreen", "darkgreen", "darkgreen"),
        lty = c(1, 2, 1, 2, 2),
        main = "ARIMA(2,1,2) Forecast (back-transformed from log scale) vs. Actual")
legend("topleft", legend = c("Train", "Actual 1960", "ARIMA Forecast", "95% CI"),
       col = c("steelblue", "black", "darkgreen", "darkgreen"), lty = c(1, 2, 1, 2))
```

:::

Hang on to these three numbers — we're comparing them directly to Step 5's in
the next step.

## Step 9: Compare Forecast Accuracy

Let's put the two forecasts side by side using the error numbers we collected
in Steps 5 and 8.

::: {.panel-tabset}

### Python

```{python}
accuracy_comparison = pd.DataFrame(
    {
        "RMSE": [rmse_decomp, rmse_arima],
        "MAE": [mae_decomp, mae_arima],
        "MAPE (%)": [mape_decomp, mape_arima],
    },
    index=["Classical Decomposition", "ARIMA(2,1,2)"],
)
print(accuracy_comparison.round(2))
```

### R

```{r}
accuracy_comparison <- data.frame(
  RMSE = c(rmse_decomp, rmse_arima),
  MAE = c(mae_decomp, mae_arima),
  MAPE = c(mape_decomp, mape_arima)
)
rownames(accuracy_comparison) <- c("Classical Decomposition", "ARIMA(2,1,2)")
print(round(accuracy_comparison, 2))
```

:::

Don't be surprised if `ARIMA(2,1,2)` doesn't clearly beat the much simpler
decomposition forecast here — and if anything, it can come out worse. That's
not a bug; it's the seasonal gap we flagged back in Step 6. Plain
`ARIMA(p, d, q)` has no seasonal term, so on a series with a strong, regular
12-month cycle like this one, it's working at a real disadvantage compared to
a method (like our decomposition forecast) that explicitly repeats the
seasonal pattern. The natural next step — adding a seasonal term to get a full
**SARIMA** model — would very likely close that gap, but it's outside the
AR/MA/ARMA/ARIMA scope of this walkthrough. Worth having ready if someone in
the room asks "why didn't ARIMA win?"

## Wrap-Up

We worked through the same nine steps twice — load, plot, decompose, test for
stationarity, forecast with decomposition, check ACF/PACF, fit AR/MA/ARMA/
ARIMA, forecast with ARIMA, and compare. A few things that stood out as we
went:

- **R**'s `ts` object carries its frequency and start date with it everywhere; in **pandas** we manage a `DatetimeIndex` ourselves, which is one extra step but also more flexible for irregular data.
- **R**'s `decompose()` and `arima()` are base functions — nothing to install beyond `tseries` for the stationarity tests. **Python** relies on `statsmodels` for essentially all of it.
- `decompose()` hands us the seasonal index directly as `$figure`; in `statsmodels` we get there by slicing the first 12 values out of the repeating `seasonal` series — same number, different route.
- `predict()` on an R `arima` fit returns log-scale predictions and standard errors we exponentiate by hand; `statsmodels`' `get_forecast().conf_int()` gives a ready-made interval, still on whatever scale we fit on.
- Plotting is base R graphics (`plot()`, `acf()`, `ts.plot()`) versus `matplotlib` — different calls, but once we know the data lines up, it's a fairly direct line-by-line translation between the two.
